---
title: "Binary AddiVortes: probit classification"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 96, out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Introduction

Binary AddiVortes is the classification variant of Stone and Gosling
(2025): the probability of the positive label is a probit function of
the sum of tessellations, fitted with the latent-variable augmentation
of Albert and Chib (1993), as BART's `pbart` fits its probit model. A
two-level factor response selects it, or `probit_outcome()` names it.
Reach for it where you would reach for `pbart`: a binary outcome whose
probability is a nonlinear function of the covariates.

## Likelihood

    y_i in {0, 1},   P(y_i = 1 | x_i) = Phi(c + f(x_i))

with Phi the standard normal distribution function, c an offset and f
the sum of tessellations. The sampler draws a latent
z_i ~ N(c + f(x_i), 1), truncated to the side its label demands, before
each sweep, and updates the ensemble as in the Gaussian model with
z - c as the response and unit variance.

## Priors

- Cell values mu ~ N(0, sigma_mu^2) with sigma_mu = 3 / (k sqrt(m)) on
  the latent scale, so the prior on f(x) puts most of its mass on
  [-3, 3] (Chipman, George and McCulloch 2010, s. 4;
  `term_params(k = 3)`).
- The cell-count, covariate-inclusion and centre priors of the Gaussian
  model, unchanged.

Fixed: the latent variance is 1, since a probit model does not identify
it, so no sigma is drawn. The offset c is `probit_outcome(offset = )`
and resolves to Phi^-1(ybar) at fit, the BART `binaryOffset` default.

## Posterior

`predict()` gives the posterior mean probability of the second level;
`interval = "credible"` an interval on the probability scale;
`type = "latent"` the per-draw latent mean c + f(x). `log_lik()` is the
Bernoulli log-likelihood, `posterior_predict()` draws labels, and
`fit$in_sample_rmse` is the root Brier score. `sigma()` is 1, a variance
ensemble is not available, and `predict(interval = "prediction")`
errors: a two-point distribution has no continuous predictive interval.

## Example

The truth is a probability surface over two covariates, steep across a
boundary oblique to both axes; labels are drawn from it.

```{r data}
library(thiessen)
library(ggplot2)

set.seed(1)
probability <- function(x) pnorm(5 * (x[, 1] + 0.3 * x[, 2] - 0.65))
draw <- function(n) {
  x <- cbind(a = runif(n), b = runif(n))
  p <- probability(x)
  list(x = x, p = p, label = factor(rbinom(n, 1, p), labels = c("no", "yes")))
}
train <- draw(300)
test <- draw(500)

fit <- thiessen(train$x, train$label, seed = 1)
fit
```

A two-level factor response becomes 0 and 1 with the first level as the
zero, as `glm()` treats one; the levels are kept on the fit.
`predict()` returns the probability of the second level, "yes":

```{r predict}
estimate <- predict(fit, test$x, interval = "credible", level = 0.9)
head(estimate, 3)
```

Against the truth on the held-out rows: the root mean squared error of
the probability, the accuracy at 0.5, and the coverage of the 90 per
cent credible interval.

```{r score}
data.frame(
  rmse = sqrt(mean((estimate[, "fit"] - test$p)^2)),
  accuracy = mean((estimate[, "fit"] > 0.5) == (test$label == "yes")),
  coverage = mean(test$p >= estimate[, "lower"] & test$p <= estimate[, "upper"])
)
```

The figure plots the estimated probability against the true one for the
held-out rows, coloured by the label that was drawn.

```{r figure}
ggplot(data.frame(truth = test$p, estimate, label = test$label),
       aes(truth, fit, colour = label)) +
  geom_abline(linetype = "dashed") +
  geom_point(alpha = 0.6) +
  labs(x = "P(y = yes | x), truth", y = "posterior mean probability")
```

## References

Albert, J. H. and Chib, S. (1993). Bayesian analysis of binary and
polychotomous response data. *Journal of the American Statistical
Association* 88(422), 669-679. doi:10.1080/01621459.1993.10476321

Chipman, H. A., George, E. I. and McCulloch, R. E. (2010). BART:
Bayesian additive regression trees. *The Annals of Applied Statistics*
4(1), 266-298. doi:10.1214/09-AOAS285

Stone, A. and Gosling, J. P. (2025). AddiVortes: (Bayesian) additive
Voronoi tessellations. *Journal of Computational and Graphical
Statistics* 34(3), 859-871. doi:10.1080/10618600.2024.2414104
