---
title: "Chains, convergence and compute"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 3, dpi = 96, out.width = "100%"
)
```

A default fit runs four chains and pools their draws, so every fit
carries the convergence diagnostics of Vehtari and others (2021):
rank-normalised split R-hat and the bulk and tail effective sample
sizes. This page shows what is monitored, what a fit that fails the
check looks like beside one that passes, what the thresholds mean, and
what the chains cost in time.

```{r data}
library(thiessen)
library(posterior)

set.seed(1)
n <- 200
x <- cbind(a = runif(n), b = runif(n))
y <- 2 * (x[, "a"] - 0.5)^2 + 0.5 * x[, "b"] + rnorm(n, sd = 0.1)
```

## What is monitored

The check runs over sigma and over the mean function at up to twenty
training rows, spread through the data. Each chain has its own seed,
derived from `seed` in the core, and each starts from the same
initial state, so the chains differ by their random streams alone. A
fit warns where R-hat exceeds 1.01 or an effective sample size falls
below 400, the thresholds Vehtari and others (2021) give, and
`print()` and `summary()` repeat the warning.

## A fit that warns, and one that does not

A schedule of 50 burn-in sweeps and 100 draws is too short for the
chains to agree:

```{r short}
short <- thiessen(x, y, thiessen_control(
  general_params = general_params(burn_in = 50, draws = 100)
), seed = 1)
```

The default schedule of 200 burn-in sweeps and 1000 draws per chain
passes on these data:

```{r long}
long <- thiessen(x, y, seed = 1)
long$convergence[c("rhat", "ess_bulk", "ess_tail")]
```

The trace of sigma over the chains shows what the schedule buys: 100
draws per chain against 1000. The short fit fails the check on the
effective sample size, 80 against 400, and on an R-hat of 1.06 over
the monitored quantities; the chains of the default fit overlap and
its smallest effective sample size is in the hundreds.

```{r trace-short}
bayesplot::mcmc_trace(as_draws_array(short), pars = "sigma")
```

```{r trace-long}
bayesplot::mcmc_trace(as_draws_array(long), pars = "sigma")
```

`summarise_draws()` gives the per-variable numbers behind the check.
Here, sigma and the mean function at three training rows:

```{r table}
draws <- as_draws_df(long)
summarise_draws(
  subset_draws(draws, variable = c("sigma", "mu[1]", "mu[2]", "mu[3]")),
  "mean", "sd", "rhat", "ess_bulk", "ess_tail"
)
```

## What to change when the fit warns

More draws per chain, `general_params(draws = )`, is the first
remedy: the effective sample size grows with the draws, and R-hat
falls as the chains have longer to mix. More chains add draws at the
same cost per draw. `thinning` reduces the stored draws without adding
information, so it is not a remedy for a low effective sample size.

The default schedule is short for a harder problem. On Friedman #1
with n = 200 and p = 10 it reaches a smallest effective sample size of
about 130 and an R-hat of about 1.02, so the fit warns; the [Gaussian
regression](gaussian.html) page runs 3000 draws per chain, which
clears both thresholds.

A fit of one chain carries no diagnostics and says so:

```{r one-chain}
one <- thiessen(x, y, seed = 1, chains = 1)
one
```

## Compute

Cost is linear in the sweeps and close to linear in the rows and in
the tessellation count. The chains run on `getOption("mc.cores", 1L)`
threads, read when the fit is called, as Stan's interfaces do: a
session that sets nothing runs the four chains on one thread and pays
four chains; `options(mc.cores = 4)` runs them on four cores for the
same draws. The draws do not depend on the thread count.

Effective sample size per second is the number to compare schedules
and thread counts by, since it measures information gained rather than
draws stored. On the machine that built this page, the default fit on
one thread and on two:

```{r timing}
seconds <- function(threads) {
  elapsed <- system.time(
    fit <- thiessen(x, y, seed = 1, threads = threads)
  )[["elapsed"]]
  c(threads = threads, seconds = round(elapsed, 1),
    min_ess = round(fit$convergence$ess_bulk),
    ess_per_second = round(fit$convergence$ess_bulk / elapsed))
}
rbind(seconds(1), seconds(2))
```

Pooling the draws and the diagnostics run on one thread after the
sweeps, so the speed-up from more threads is less than the thread
count; at n = 200, four threads take about 45 per cent of the
one-thread time. `predict()` splits its rows over the fit's thread
count, or over its own `threads` argument.

Progress over the whole fit is signalled with progressr, so nothing is
printed unless a session chooses a handler; the [getting
started](thiessen.html) page shows the two lines that set one.

## References

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B. and Buerkner, P.-C.
(2021). Rank-normalization, folding, and localization: an improved R-hat
for assessing convergence of MCMC. *Bayesian Analysis* 16(2), 667-718.
doi:10.1214/20-BA1221
