---
title: "Gaussian regression"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 96, out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Introduction

The Gaussian model is AddiVortes as Stone and Gosling (2025) state it:
a continuous response, a mean function that is a sum of Voronoi
tessellations, and one residual variance. It is the model
`thiessen()` fits to a numeric response when nothing else is named, and
the default configuration is the published one. Reach for it where you
would reach for BART's `wbart`: a nonlinear regression with
interactions you do not want to specify, on tens to thousands of rows.

## Likelihood

    y_i = f(x_i) + e_i,   e_i ~ N(0, sigma^2),   i = 1, ..., n

with f(x) the sum of m tessellations, each a piecewise-constant
function on a Voronoi partition of a random subset of the covariates.
The [model description](model-description.html) page states f and the
partition in full.

## Priors

- Cell values mu ~ N(0, sigma_mu^2) with sigma_mu = 0.5 / (k sqrt(m))
  on the response scaled to [-0.5, 0.5], so the prior on f(x) puts most
  of its mass across the range of the data (`term_params(k = 3)`).
- Residual variance sigma^2 ~ nu lambda / chi^2_nu, with lambda set so
  that Pr(sigma < sigma_hat) = q for sigma_hat the residual standard
  deviation of a least-squares fit (`gaussian_outcome(nu = 6, q =
  0.85)`).
- Cells per tessellation b - 1 ~ Poisson(lambda_c)
  (`term_params(lambda_c = 5)`); active covariates d - 1 ~
  Binomial(p - 1, omega / p) (`structure_params(omega = min(3, p))`);
  centre coordinates N(0, sigma_c^2) on the scaled covariates
  (`geometry_params(sigma_c = 0.8)`).

Nothing is fixed: every quantity above is sampled or set by the data.
The [priors](priors.html) page draws from each of them.

## Posterior

A fit carries the kept draws of every chain. `predict()` gives the
posterior mean of f(x), a credible interval for it, or a predictive
interval for a new observation; `predict(type = "draws")` the per-draw
mean function and `type = "variance"` the per-draw sigma^2; `sigma()`
the posterior mean of sigma; `log_lik()` the pointwise log-likelihood
under N(f_d(x), sigma_d^2) for each draw d; and
`posterior::as_draws_df()` the draws as variables `mu[i]`, `sigma`,
`cell_count` and `dimension_count`.

## Example

Friedman #1 (Friedman 1991) is the test function the BART papers use:
five informative covariates, five noise, and noise of standard
deviation 1. The truth is known, so the held-out error is measured
against the noise-free function and not against a noisy response.

```{r data}
library(thiessen)
library(ggplot2)

friedman <- function(n, p = 10, sd = 1) {
  x <- matrix(runif(n * p), n, p, dimnames = list(NULL, paste0("x", 1:p)))
  f <- 10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 0.5)^2 +
    10 * x[, 4] + 5 * x[, 5]
  list(x = x, f = f, y = f + rnorm(n, sd = sd))
}

set.seed(1)
train <- friedman(200)
test <- friedman(500)
```

The default schedule of 1000 draws per chain is short for this problem:
with p = 10 the effective sample size of the mean function falls below
the threshold of 400 and the fit warns. Three thousand draws per chain
clear it, so that is the one setting changed from the defaults.

```{r fit}
control <- thiessen_control(general_params = general_params(draws = 3000))
fit <- thiessen(train$x, train$y, control, seed = 1)
fit
```

`sigma()` recovers the noise scale the data were made with:

```{r sigma}
sigma(fit)
```

On the held-out rows, the root mean squared error of the posterior mean
against the noise-free truth, the coverage of the 90 per cent credible
interval for f and of the 90 per cent predictive interval for y, and
the mean width of each:

```{r score}
credible <- predict(fit, test$x, interval = "credible", level = 0.9)
predictive <- predict(fit, test$x, interval = "prediction", level = 0.9)

data.frame(
  rmse = sqrt(mean((credible[, "fit"] - test$f)^2)),
  credible_coverage = mean(test$f >= credible[, "lower"] &
                             test$f <= credible[, "upper"]),
  credible_width = mean(credible[, "upper"] - credible[, "lower"]),
  predictive_coverage = mean(test$y >= predictive[, "lower"] &
                               test$y <= predictive[, "upper"]),
  predictive_width = mean(predictive[, "upper"] - predictive[, "lower"])
)
```

The figure plots the posterior mean against the truth for sixty of the
held-out rows, with the 90 per cent credible interval as the bar.

```{r figure}
shown <- sample(nrow(test$x), 60)
ggplot(data.frame(truth = test$f[shown], credible[shown, ]),
       aes(truth, fit)) +
  geom_abline(linetype = "dashed") +
  geom_pointrange(aes(ymin = lower, ymax = upper), colour = "steelblue",
                  size = 0.3) +
  labs(x = "f(x), noise-free truth", y = "posterior mean of f(x)")
```

`log_lik()` at new data gives the log predictive density of each
held-out observation, averaged over the draws; its mean over rows is
the log score.

```{r log-score}
log_likelihood <- log_lik(fit, newdata = test$x, y = test$y)
mean(log(colMeans(exp(log_likelihood))))
```

`variable_inclusion()` reports where the ensemble spent its
dimensions. With p = 10 and `omega` at its default of 3, each
tessellation holds about three covariates, and the shares separate
only weakly from the uniform 0.1: the ensemble reaches the five
informative covariates through many small tessellations rather than
by concentrating on them. Read the shares as where the dimensions
went, not as variable selection:

```{r inclusion}
round(variable_inclusion(fit), 3)
```

## References

Chipman, H. A., George, E. I. and McCulloch, R. E. (2010). BART:
Bayesian additive regression trees. *The Annals of Applied Statistics*
4(1), 266-298. doi:10.1214/09-AOAS285

Friedman, J. H. (1991). Multivariate adaptive regression splines. *The
Annals of Statistics* 19(1), 1-67. doi:10.1214/aos/1176347963

Stone, A. and Gosling, J. P. (2025). AddiVortes: (Bayesian) additive
Voronoi tessellations. *Journal of Computational and Graphical
Statistics* 34(3), 859-871. doi:10.1080/10618600.2024.2414104
