---
title: "H-AddiVortes: heteroscedastic regression"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 96, out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Introduction

H-AddiVortes is the heteroscedastic variant of Stone and Gosling
(2025): the noise variance is itself a function of x, a product of
variance tessellations, with the structure of HBART (Pratola, Chipman,
George and McCulloch, 2020). It is selected by attaching a variance
ensemble to the Gaussian family, a positive tessellation count on
`variance_params`; the paper's count is 40. Reach for it where the
spread of the response changes across the covariate space and a
predictive interval of one width would be wrong somewhere.

## Likelihood

    y_i = f(x_i) + s(x_i) e_i,   e_i ~ N(0, 1)
    s^2(x) = prod_{j=1}^{m'} v(x; T'_j, V_j)

with f the sum of m mean tessellations and s^2 the product of m'
variance tessellations, v(x; T', V) the value of the cell of T' that x
falls in. One sweep updates the variance ensemble given the residuals
y - f, then the mean ensemble with per-observation precision
1 / s^2(x_i).

## Priors

- Mean cells as in the Gaussian model, mu ~ N(0, sigma_mu^2) with
  sigma_mu = 0.5 / (k sqrt(m)).
- Each variance cell v ~ inverse gamma(nu' / 2, nu' lambda' / 2), with nu'
  and lambda' set from `gaussian_outcome(nu, q)` so that the prior mean
  of s^2(x) equals the prior mean of the Gaussian model's sigma^2
  whatever m' is (HBART s. 3.2). This needs nu > 2.
- Both ensembles share the cell-count, covariate-inclusion and centre
  priors: `geometry` and `structure` set on `mean_params` apply to
  `variance_params` too.

## Posterior

`predict()` gives the posterior mean of f(x) and `interval =
"prediction"` an interval whose width follows s(x);
`type = "variance"` gives s^2(x) per draw, the square of HBART's
`sdraws`. `sigma()` is not defined, since there is no single residual
scale, and the family must be Gaussian: the probit family fixes its
latent scale at 1, so it takes no variance ensemble.

## Example

The mean is linear in the first covariate and the noise scale grows
with it, from 0.05 to 0.35.

```{r data}
library(thiessen)
library(ggplot2)

set.seed(1)
n <- 300
x <- cbind(a = runif(n), b = runif(n))
scale <- function(x) 0.05 + 0.3 * x[, "a"]
y <- x[, "a"] + rnorm(n, sd = scale(x))

control <- thiessen_control(
  mean_params = term_params(tessellations = 200),
  variance_params = term_params(tessellations = 40),
  general_params = general_params(draws = 2000)
)
fit <- thiessen(x, y, control, seed = 1)
fit
```

The posterior mean of s(x) along `a`, with `b` fixed, against the scale
the data were made with:

```{r figure}
grid <- cbind(a = seq(0, 1, length.out = 100), b = 0.5)
estimate <- sqrt(colMeans(predict(fit, grid, type = "variance")))

ggplot(data.frame(a = grid[, "a"], estimate, truth = scale(grid)), aes(a)) +
  geom_line(aes(y = estimate), colour = "steelblue") +
  geom_line(aes(y = truth), linetype = "dashed") +
  labs(y = "s(x), posterior mean")
```

Predictive intervals widen where the variance is high, which is the
point of the model. The mean width of the 90 per cent interval over the
fifty rows with the smallest `a` and the fifty with the largest:

```{r widths}
interval <- predict(fit, x, interval = "prediction", level = 0.9)
width <- interval[, "upper"] - interval[, "lower"]
order <- order(x[, "a"])
c(low_a = mean(width[head(order, 50)]), high_a = mean(width[tail(order, 50)]))
```

## References

Pratola, M. T., Chipman, H. A., George, E. I. and McCulloch, R. E.
(2020). Heteroscedastic BART via multiplicative regression trees.
*Journal of Computational and Graphical Statistics* 29(2), 405-417.
doi:10.1080/10618600.2019.1677243

Stone, A. and Gosling, J. P. (2025). AddiVortes: (Bayesian) additive
Voronoi tessellations. *Journal of Computational and Graphical
Statistics* 34(3), 859-871. doi:10.1080/10618600.2024.2414104
